\thetitle

### 2021-03-01 writing/challenges.tex
re-organize and re-write secs. 1 thru 6

The {\em first derivative challenge\/} is to find a representative collection of
impedance mismatches. They must represent mistakes that programmers accidentally
create and that the run-time checks of academic systems catch. In other words,
the experiment calls for a collection of type mistakes in mix-typed programs
that is representative of those ``in the wild.''  Unfortunately no such
collection exists, and with good reason. The kind of mistakes needed are
typically detected by unit or integration tests; even if it takes some time to
find their sources, these mistakes do not make it into code repositories with
appropriate commit messages.

@@

@@

The {\em second derivative challenge\/} concerns the differences among the blame
assignment of the three semantic variants.  While Natural assigns blame to {\em
one\/} component, Transient assigns blame to a sequence of components. The
Erasure semantics does not blame a component {\it per se\/}, but it comes with
an exception location and a stack trace, which implicitly suggests parties to
blame.  The different strategies greatly affect the behavior of the rational
programmer (and real programmers, too).

@@

Overcoming these challenges yields a systematic approach: 
\begin{itemize}

\item interesting mutators generate interesting type-level errors and thus
debugging scenarios (section~\ref{sec:mutate});

\item an algorithm simulates the modes of the rational programmer on these
scenarios (section~\ref{sec:rational}); and

\item a careful sampling of the space yields experimental
data about the bug finding process (section~\ref{sec:sample}).
\end{itemize}
Getting all three aspects right is necessary to obtain a computationally
feasible and statistically sound experiment.

{\bf Note} A coincidental challenge is the disparity of gradually typed
languages. While the three approaches to blame are well understood in theory,
they have been implemented for rather different untyped languages and with
vastly different type systems. To eliminate the effect of these differences, it
is necessary to implement all three approaches in a single language so that an
application of the evaluation method can use the same test programs, the same
type system, and hence the same debugging scenarios.

The results in this paper are for a system that builds on Greenman and
Felleisen's effort of comparing the performance of Natural and Transient in
Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
Racket and comes with a Transient semantics that faithfully implements the
design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
Erasure semantics drops the type information and runs the programs as if they
were plain Racket programs. The next section explains this system in detail.

>>>>

The rational programmer works like an algorithm. Implementing it demands a large
number of representative debugging scenarios; a reconciliation of three
different kinds of error messages; and a proper sampling strategy for the huge
space of possibilities. A coincidental challenge is the disparity of gradually
typed languages. To eliminate this variable from the comparison, this paper uses
Racket, which is thus far the only language in which all three major soundness
mechanisms are available in a reasonably robust and directly comparable manner;
for the implementation of the transient semantics, see the parallel submission
of an experience report~\cite{ttt21}.

The {\em first challenge\/} is to find a representative collection of impedance
mismatches. They must represent mistakes that programmers accidentally create
and that the run-time checks of academic systems catch. In other words, the
experiment calls for a collection of type mistakes in mix-typed programs that is
representative of those ``in the wild.''  Unfortunately no such collection
exists, and with good reason. The kind of mistakes needed are typically detected
by unit or integration tests; even if it takes some time to find their sources,
these mistakes do not make it into code repositories with appropriate commit
messages.

@@

Based on a related experience, Gopinath and Walkingshaw~\cite{gw-mutation} write,
``existing mutation frameworks \ldots\ do not generate the kinds of mutations
needed to best evaluate type annotations'' and, worse, ``it is surprisingly
difficult to come up with mutants that actually describe subtle type faults.''
While the goal of their work---to evaluate the quality of types in
Python---is unrelated to the one here, the mechanism is related. And their
judgment confirms the experience of the authors.

@@

The {\em second challenge\/} concerns the differences among the blame assignment
of the three semantic variants.  While Natural assigns blame to {\em one\/}
component, Transient assigns blame to a sequence of components. The Erasure
semantics does not blame a component {\it per se\/}, but it comes with an
exception location and a stack trace, which implicitly suggests parties to
blame.  The different strategies greatly affect the behavior of the rational
programmer (and real programmers, too).

@@

The next three sections explain how to overcome the three challenges
in detail.

% Overcoming these challenges yields a systematic approach: 
% \begin{itemize}

% \item interesting mutators generate interesting type-level errors and thus
% debugging scenarios (section~\ref{sec:mutate});

% \item an algorithm simulates the modes of the rational programmer on these
% scenarios (section~\ref{sec:rational}); and

% \item a careful sampling of the space yields experimental
% data about the bug finding process (section~\ref{sec:sample}).
% \end{itemize}
% Getting all three aspects right is necessary to obtain a computationally
% feasible and statistically sound experiment.

% {\bf Note} A coincidental challenge is the disparity of gradually typed
% languages. While the three approaches to blame are well understood in theory,
% they have been implemented for rather different untyped languages and with
% vastly different type systems. To eliminate the effect of these differences, it
% is necessary to implement all three approaches in a single language so that an
% application of the evaluation method can use the same test programs, the same
% type system, and hence the same debugging scenarios.

% The results in this paper are for a system that builds on Greenman and
% Felleisen's effort of comparing the performance of Natural and Transient in
% Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
% Racket and comes with a Transient semantics that faithfully implements the
% design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
% Erasure semantics drops the type information and runs the programs as if they
% were plain Racket programs. The next section explains this system in detail.

### 2021-03-01 writing/introduction.tex
re-organize and re-write secs. 1 thru 6

Industrial implementors of gradual typing systems have almost completely ignored
blame assignment.  Systems such as Flow,\footnote{\url{https://flow.org}}
Hack,\footnote{\url{https://hacklang.org}} or
TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE
actions and for finding typos in code. Then their compilers remove types and
rely on the built-in safety checks of the underlying language to catch any
problems.

@@

This paper's first contribution is {\em an automated method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see section~\ref{sec:why-rational}). In the case
of an impedance mismatch, the available information consists of the error
message and the current state of the program. The rational programmer can hence
use the former to change the latter---and this systematic, information-driven
process can be implemented.\footnote{\citet{cc-snapl-19} use this tracing
method to localize type errors reported by gradual typing systems based on
inference.} Implementing it means overcoming three major challenges: injecting
representative impedance mismatches; putting the various kinds of error
information to comparable use; and sampling the huge space of possibilities
properly (see section~\ref{sec:challenges}).

% The method simulates a programmer who follows the Wadler-Findler slogan quoted
% above; its automation borrows elements of of Lazarek et
% al.~\cite{lksfd-popl-2020}'s recent work.  Such a \emph{rational programmer} can find impedance-mismatch
% bugs by adding types to a partially typed program (see
% section~\ref{sec:landscape}). That is, when the run-time system signals an
% impedance mismatch, a good blame message helps the programmer infer where to
% add an additional type specification because, by the basic theory, well-typed
% components cannot cause impedance mismatches. Otherwise, if the additional type
% does {\em not\/} cause a static type error, the program is run again to signal
% another violation. Then the programmer can heed blame one more time to add yet
% another type specification. Evaluating the quality of blame assignment thus
% becomes a question of whether the tracing the error messages finds the source
% of the mistake and how many steps it takes to get to this point.

The paper's second contribution is a set of {\em results from applying the
evaluation method to three distinct blame assignment strategies and
approximately 72,200 cases\/} (see section~\ref{sec:results}): (1) {\it
Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries; (2) {\it
Natural\/}, i.e. the use of higher-order contract system and its blame
assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
industrial systems, which forego blame assignment and rely on error messages
from the safety checks of the underlying programming language.

The results of the method's first application (see section~\ref{sec:discussion})
are at least somewhat surprising. In principle they validate the conjectures
behind the work of theoreticians. First, a good blame assignment strategy helps
with the search for impedance mismatches between the specified types and the
behavior of untyped components.  Second, Natural's wrapper-based blame tracking
is more useful than Transient's ``collective blame'' tracking algorithm, which
in turn is superior to Erasure. {\em But\/}, the application of the method also
indicates problems with the expectations of theoreticians. The first problem is
that the cost of Transient's blame can be huge. The second one concerns the
difference between the methods; neither is {\it Natural\/} vastly better than
{\it Transient\/} nor are the two blame assignment methods clearly preferable to
{\it Erasure\/}.  In turn, these problems suggests that, on one hand, the
existing theory does not predict practice properly, and on the other hand, the
existing practice may need to find additional ways to collect data.

>>>>

Industrial implementors of gradual typing systems have almost
completely ignored blame assignment.  Systems such as Flow, Hack, or
TypeScript\footnote{See \url{https://flow.org},
\url{https://hacklang.org}, and \url{https://www.typescriptlang.org},
respectively.} exploit types for IDE actions and for finding typos in
code. Then their compilers remove types and rely on the built-in
safety checks of the underlying language to catch any problems.

@@

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see
sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be
implemented.\footnote{\citet{cc-snapl-19} use this tracing method to localize
type errors in inference-based gradual typing systems.}  Implementing it means
overcoming major challenges: injecting representative impedance mismatches;
putting the various kinds of error information to comparable use; and sampling
the huge space of possibilities (see sec.~\ref{sec:challenges}).

The paper's {\it second contribution\/} is a set of {\em results from applying
the evaluation method to three distinct blame assignment strategies and
approximately 72,200 cases\/} (see sec.~\ref{sec:results}): (1) {\it
Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries; (2) {\it
Natural\/}, i.e. the use of higher-order contract system and its blame
assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
industrial systems, which forgo blame in favor of error messages from the safety
checks of the underlying language.---The results (see
sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
they validate the conjectures behind the work of theoreticians.  First, a good
blame assignment strategy helps with the search for impedance mismatches between
the specified types and the behavior of untyped components.  Second, Natural's
wrapper-based blame tracking is more useful than Transient's ``collective
blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
the application of the method also indicates problems with the expectations of
theoreticians. The first problem is that the cost of Transient's blame can be
huge. The second one concerns the difference between the methods; neither is
{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
problems suggests that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need to find
additional ways to collect data.

### 2021-03-01 writing/landscape.tex
re-organize and re-write secs. 1 thru 6

%% -----------------------------------------------------------------------------
\def\fname#1#2{$\underline{\hbox to #2in{#1}}$\\[-2ex]}

\begin{figure*}[hbt]\footnotesize
  \vspace{-1mm}

\begin{minipage}[t]{2.2in}
\begin{alltt}\fname{untyped/pack-lib}{1.8}
(provide pack _ _ _)

_ _ _  dependencies     _ _ _
_ _ _  and definitions  _ _ _

(define (pack d)
  ;; process JSON data and 
  ;; package it in a dictionary
  ;; (association list)
  (make-hash _ _ _) ;; BUG!)

\end{alltt}
\end{minipage}\begin{minipage}[t]{2.2in}
\begin{alltt}\fname{typed/pack-lib}{1.7}
(provide typed-pack)

(require/typed pack-lib
  [pack 
  (-> JSON 
      (Listof 
       (Pairof Symbol String)))])

(define typed-pack pack) 
\end{alltt}
\smallskip
\begin{alltt}\fname{typed/crypto-pack-lib}{1.7}
(provide crypto-pack)

(require/typed pack-lib
 [pack 
  (-> JSON 
      (Listof 
       (Pairof Symbol String)))])

(define (crypto-pack [d : JSON])
  (pack (encrypt d _ _ _)))

 \end{alltt}
\end{minipage}\begin{minipage}[t]{2.2in}
\begin{alltt}\fname{untyped/client}{1.65}
(require json 
         typed/pack-lib
           typed/crypto-pack-lib)

_ _ _  other dependencies  _ _ _
_ _ _  and definitions     _ _ _

;; read data from files, pack 
;; and share securely

(define public-data 
  (typed-pack 
   (read-json 
    "public-records")))

(define secret-data 
  (crypto-pack 
   (read-json 
    "medical-records")))

 _ _ _ rest of client _ _ _
\end{alltt}\end{minipage}

   \caption{One mix-typed program, three interpretations} \label{fig:example}

  \vspace{-1mm}
\end{figure*}
%% -----------------------------------------------------------------------------

The program sketch in figure~\ref{fig:example} demonstrates the
differences between Natural, Transient and Erasure using Typed Racket
syntax.  The modules on the left and right are untyped while the two
in the middle are typed.

The \texttt{untyped/pack-lib} module (left) provides, among others, a function
\texttt{pack} that consumes JSON data and packages it in a dictionary. However,
by mistake \texttt{pack} returns a hash table instead of the association list
its documentation promises.  The \texttt{client} module (right) uses
\texttt{pack} indirectly to package public and sensitive data it reads from the
filesystem. It does so through two different intermediary typed modules
(center). The first, {\tt typed/pack-lib}, imports \texttt{pack} and re-exports
it as \texttt{typed-pack} asserting that it is a function that consumes {\tt
JSON} and returns a list associating {\tt Symbol}s with {\tt String}s. The
second typed module, {\tt typed/crypto-pack-lib}, also imports \texttt{pack} and
assumes for it the same type as {\tt typed/pack-lib}. Instead of exporting it,
it uses the function in the exported {\tt crypto-pack} function, which encrypts
its input before passing it to \texttt{pack}.

Under the {\it Natural\/} semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017},
functions imported into and exported from typed modules are wrapped in proxies
that enforce the static type discipline with run-time checks and track
responsibilities. Thus, when \texttt{pack} is imported into a typed module, the
run-time system checks that it is a function and wraps it in a protective proxy,
which in turn, enforces the type of the function with run-time checks.  In
essence, this check protects typed modules when they call an untyped function
because it may not produce the promised kind of value, as is the case here with
{\tt pack} from {\tt untyped/pack-lib}. Analogously the run-time system wraps
exported functions such as {\tt crypto-pack} of a typed modules in a proxy that
checks its arguments.  This check protects functions exported from type modules
against applications in untyped code that supply the wrong kind of argument
value.

As this analysis implies, if the return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt
untyped/pack-lib}. If the argument-type check fails, responsibility lies with
{\tt client}. Due to the proxies wrapped around functions, Natural can easily
track the responsible party for each check. Thus, in the example of
figure~\ref{fig:example}, as \texttt{typed-pack} returns, the return-type check
fails, and Natural blames \texttt{pack-lib}.

Under the Transient semantics~\cite{vss-popl-2017}, typed code is compiled so
that all entry points to functions check their arguments at run time and all
function calls check their return values against the expected type.
Furthermore, Transient uses \emph{shallow} checks, meaning they check only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the innards of a
complex value are checked on a piecemeal basis.

As a result, the call to \texttt{typed-pack} does not signal an error because
the call takes place in the untyped {\tt client} module, which is compiled in
the usual manner. Because {\tt pack} is called in the {\tt
typed/crypto-pack-lib} module, Transient's inlined checks make sure that the
imported \texttt{pack} is a function and that the result of its call in is a
list. This last check fails when \texttt{client} calls to {\tt crypto-pack}.

In order to pinpoint the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed/untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to {\tt typed/pack-lib} and
from {\tt typed/pack-lib} to {\tt client} with the type that appears in the {\tt
required/typed} forms in the example. Since the failed check corresponds to the
return type of \texttt{pack}, Transient blames the sources of the two boundary
crosses, which in this particular case is \texttt{untyped/pack-lib} for both. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

Under the Erasure semantics, the compiler checks the specified types and then
discards them when it generates code. The generated code includes run-time
checks that ensure the dynamic safety of all operations as specified in the
underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
call to {\tt crypto-pack} signals an error due to the gradual type system.

When run, the program may raise an exception if \texttt{client} tries to
inspect the elements of the list that \texttt{typed-pack} and
\texttt{crypto-pack} are supposed to produce. The information in this exception,
plus its stack trace, or may not help the programmer find the source of the
impedance mismatch between the specified return types of {\tt pack} in the two
typed modules and the actual return value of {\tt pack}.

All three systems used for the experiment are variants of Racket. The
use the Typed Racket syntax and type checker for all three
semantics. The Natural semantics is inherited as is, while the Erasure
semantics is the one of plain Racket.  The Transient system implements
the design of~\citet{vss-popl-2017}, though constructing it demands
two design decisions: which shallow checks to use and where, plus how
to maintain a blame map in a full-fledged run-time system.

For certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient can
realize shallow checks with the predicate from Racket.  Other types require
custom predicates; for example the type {\tt Index} of valid array indices
allows a range of positive exact integers.  Many custom predicates run in
constant time, but some, such as the one for {\tt (Union T1 ... Tn)}, run in
linear time. The key point is that no shallow check inspects beyond the top
constructor of each type.

Once the type checker has completed its work, Transient rewrites the
source to insert run-time checks. Technically speaking, the rewriting
pass wraps all elimination in applications of shallow checks.  For
example, if the variable {\tt f} has the static type {\tt (-> String
String)}, the application {\tt (f x)} must be rewritten to check for a
string result because {\tt f} could be an untyped function.

The rewriting pass also inserts hooks for updating the blame map.  When a value
crosses a module boundary, such a hook records a type assumption in the blame
map.  The in-lined checks for elimination forms extend the blame map with
further information. For instance, the check of the result of {\tt (f x)} adds
an entry to the blame map that records the expected type of the result as the
return type of the type of {\tt f}.  If a Transient check ever fails, the
run-time checking system of Transient Racket walks over the information in the
blame map and reports a sequence of blamed components.

%% MF: I have no clue what the following contributes:

% \smallskip
% \noindent
% {\bf Note.} As the common basis for the three semantics, we choose Typed
% Racket's migratory typing. In contrast to the initial formulation of gradual
% typing~\cite{st-sfp-2006}, migratory typing does not provide support for type
% \texttt{Dyn}, components are either untyped or fully typed.
% %% MF: does it ever:
% %% and the gradual type system does not affect the compilation of untyped components. 
% Despite, these differences, Typed Racket is the most
% mature system for sound mix-typed code
% and recent theoretical work has established that
% its migratory typing offers a setting for comparing different approaches to gradual
% typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.

>>>>

%% -----------------------------------------------------------------------------

The three blame strategies rely on three different ways of catching problems
with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
Erasure\/}. This section summarizes these options with one illustrative example
using Racket syntax. The informed reader may wish to merely scan the
self-contained section.

Consider the program sketch in figure~\ref{fig:example}. Each box represents a
module: the top bar lists the name and whether it is using typed (\typecolor) or
untyped (\dyncolor) syntax:
\begin{description}

\item[\texttt{pack-lib}] (at the top right) represents a library that provides,
among others, a function {\tt pack}. As the documentation says, this function
consumes JSON data and packages it in a dictionary. by mistake, though, the
function returns a hash table instead of the association list its documentation
promises.

\item[\texttt{types}] (at the top left) is one of three modules that overlays
types onto this library. This specific modules defines types in common to the
two other typed libraries.

\item[\texttt{typed-pack-lib}] (at the mid-level on the left), imports {\tt pack}
and re-exports it as \texttt{typed-pack} asserting that it is a function that
consumes {\tt JSON} and returns a list associating {\tt Symbol}s with {\tt
String}s. In other words, it formalizes the comments in {\tt pack-lib}.

\item[\texttt{crypto-pkac-lib}] (at the bottom left) also imports \texttt{pack}
and assumes for it the same type as {\tt typed-pack-lib}. It applies the
function in the definition of the exported {\tt crypto-pack} function, which
encrypts its input before passing it to \texttt{pack}.

\item[\texttt{client}] (at the bottom right) uses {\tt pack}
indirectly. Specifically, it goes through the two intermediary typed modules to
import them. Imagine a programmer who relies on the types in the \typecolor\
modules as checked documentation but prototypes the client in an untyped manner.

\end{description}
Each of the three semantics assigns a different outcome to the client module and
signals different kinds of errors.

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\cite{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function with run-time checks.
In essence, this check protects typed modules when they call an untyped function
because it may not produce the promised kind of value, as is the case here with
{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps exported
functions such as {\tt crypto-pack} of a typed modules in a proxy that checks
its arguments.  This check protects functions exported from type modules against
applications in untyped code that supply the wrong kind of argument value.

As this analysis implies, if the return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
the argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{typed-pack} returns, the return-type check fails, and Natural blames
\texttt{typed-pack-lib}.

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\cite{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the innards of a
complex value are checked on a piecemeal basis.

As a result, the call to \texttt{typed-pack} does {\em not\/} signal an error
because it takes place in the untyped {\tt client} module, which is compiled in
the usual manner. Because {\tt pack} is called in the {\tt crypto-pack-lib}
module, Transient's inlined checks make sure that the imported
\texttt{pack} is a function and that the result of its call in is a list. This
last check fails when \texttt{client} calls to {\tt crypto-pack}.

In order to pinpoint the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed/untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

Under {\it the Erasure semantics\/}, the compiler checks the specified types and
then discards them when it generates code. The generated code includes run-time
checks that ensure the dynamic safety of all operations as specified in the
underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
call to {\tt crypto-pack} signals an error due to the gradual type system. When,
at some later point, {\tt client} tries to inspect the elements of the lists
that \texttt{typed-pack} and {\tt crypto-pack} are supposed to produce, Racket's
safety check signal a violation and point to some place in {\tt client}. The
information in this exception, plus its stack trace, may help the programmer
find the source of the impedance mismatch between the specified return types of
{\tt pack} in the two typed modules and the actual return value of {\tt pack}.

The following table summarizes the illustration:
%% -----------------------------------------------------------------------------
\begin{center}
\begin{tabular}{l|ll}
                                 \multicolumn{3}{c}{the evaluation of} \\
\relax                  & {\tt public-data}             & {\tt secret-data} in {\tt client}, \\
{under the $\cdot$ semantics yields} & & \\ \hline
%% ----------------------------------------------------------------------------------------
\qquad {\it Natural\/}   & error, blaming              & error, blaming \\
     		   	 & \quad {\tt typed-pack-lib}  & \quad {\tt typed-pack-lib} \\
\qquad {\it Transient\/} & no error   		       & error, blaming the {\tt pack-lib}/ \\
      		  	 &    			       & \quad {\tt crypto-pack-lib} boundary \\
\qquad {\it Erasure\/}   & no error 		       & no error \\
\end{tabular}
\end{center}
%% -----------------------------------------------------------------------------

### 2021-03-01 writing/mutate-mutators.tex
proper voice and style
re-organize and re-write secs. 1 thru 6

Generating debugging scenarios from the benchmarks demands the design
of novel mutators. Generally speaking, a
mutator~\cite{jia2011analysis,demillo1978hints,lipton1971fault}
performs a localized syntactic change to a code basis. For the
evaluation of a blame strategy, the mutator must produce type-level
mistakes that the run-time checks of gradual typing systems or the
safety checks of the underlying language can detect. Once detected,
the rational programmer can attempt to detect the mistake by gradually
adding types to blamed modules.

Figure~\ref{table:mutation-ops} describes the sixteen mutators used
for the data reported here. They are inspired by the authors
decades-long experience of making type-level mistakes in Typed Racket,
some of which take a non-trivial effort to debug.  For each mutaror,
the figure provides a short description and an example mutation.

Most of these mutators are self-explanatory.  The first twelve apply
to most gradually typed languages, including those with classes.  The
last four target distinguishing features of Typed Racket.
Specifically, {\tt arithmetic} may replace a {\tt +} with a {\tt -} in
an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, {\tt +}'s result is a {\tt
Positive-Integer} when all arguments are positive
integers~\cite{stathff-padl-12}. However, the result of {\tt -}
is {\tt Integer}. In the same spirit, the {\tt boolean} mutator aims
to take advantage of Typed Racket faithfulness to Racket's
``falsiness.'' Finally, {\tt negate-cond} and {\tt force-cond} attempt
to confuse occurrence typing.

>>>>

% -----------------------------------------------------------------------------

A {\it mutator\/} performs a localized syntactic change to a code base. The
result is a {\em mutant\/}.

For the evaluation of a blame strategy, the mutator must produce type-level
mistakes that the run-time checks of gradual typing systems or the safety
checks of the underlying language can detect. Once detected, the rational
programmer should be able to detect the mistake by gradually adding types to
blamed modules.

Figure~\ref{table:mutation-ops} describes the 16 mutators used for the
experiment.  They are inspired by the authors decades-long experience
of making type-level mistakes in Typed Racket, some of which take a
non-trivial effort to debug.  For each mutaror, the figure provides a
short description and an example mutation.

Most mutators are self-explanatory.  The first twelve apply to most
gradually typed languages, including those with classes.  The last
four target distinguishing features of Typed Racket.  Specifically,
{\tt arithmetic} may replace a {\tt +} with a {\tt -} in an attempt to
change the type of the result of the arithmetic operation. In Typed
Racket, {\tt +}'s result is a {\tt Positive-Integer} when all
arguments are positive integers~\cite{stathff-padl-12}. However, the
result of {\tt -} is {\tt Integer}. In the same spirit, the {\tt
boolean} mutator aims to take advantage of Typed Racket faithfulness
to Racket's ``falsiness.'' Finally, {\tt negate-cond} and {\tt
force-cond} attempt to confuse occurrence typing.

### 2021-03-01 writing/mutate.tex
re-organize and re-write secs. 1 thru 6

As section~\ref{sec:challenges} explains, the first challenge is to
generate a representative mistakes in representative software. The
following two subsections explain in reverse order.

@@

\sub{benchmarks} {Which Software to Use As Benchmarks}
\sub{interesting}{Are These Mutators Interesting?}

>>>>

Putting a rational programmer to work means generating many mutants and turning
those into debugging scenarios. Since existing mutators do not generate useful
mutants, the first step is to develop new mutators (see sec.~\ref{sub:mutate-mutators})
and to validate their suitability for this application (see sec.~\ref{sub:mutate-interesting}).

% As section~\ref{sec:challenges} explains, the first challenge is to
% generate a representative mistakes in representative software. This
% section explains the specific choices made for the first application
% of the method: the source and nature of the chosen benchmarks
% (sec.~\ref{sub:mutate-benchmarks}), the newly developed mutators
% (sec.~\ref{sub:mutate-mutators}), and why they are suitable for this
% experiment (sec.~\ref{sub:mutate-interesting}).

@@

\input{mutate-mutators-figure.tex}
\sub{mutators}   {How to Mutate Software} 
\sub{interesting}{Are These Mutators Interesting}

### 2021-03-01 writing/paper.tex
Switch to icfp format
revise intro to point to new 'why rational' section
re-organize and re-write secs. 1 thru 6

\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

\usepackage{alltt}
\usepackage{wrapfig}

@@

\sec{introduction}{Does Blame Matter}
\sec{challenges}  {Why Is It Hard to Evaluate Blame}
\sec{landscape}   {How to Think About All Three Systems}
\sec{mutate}      {How to Make Lots of Mistakes}
\sec{rational}    {How to Test the Rational Programmer}
\sec{sample}      {How to Sample the Experimental Space}
\sec{results}     {What We Discovered}
\sec{discussion}  {What We Learned}
\sec{related}     {Who has Worked on this Problem Before}
\sec{conclusion}  {What to Do Next}
%% -----------------------------------------------------------------------------

>>>>

\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

\usepackage{alltt}
\usepackage{tcolorbox}
\usepackage{wrapfig}

@@

\input{landscape-program}
\sec{introduction}{Does Blame Matter}
\sec{landscape}   {How to Think About Three Blame Systems}
\sec{why-rational}{What Is a Rational Programmer}
\sec{challenges}  {Why It Is Hard to Implement the Rational Programmer}
\sec{mutate}      {How to Make Lots of Mistakes}
\sec{rational}    {How to Make Comparable Rational Programmers}
\sec{sample}      {How to Make the Experimental Space}
\sec{results}     {What We Discovered}
\sec{discussion}  {What We Learned}
\sec{related}     {Who has Worked on this Problem Before}
\sec{conclusion}  {What to Do Next}
%% -----------------------------------------------------------------------------

### 2021-03-01 writing/results.tex
re-organize and re-write secs. 1 thru 6

We run the experimental process from section~\ref{sec:rational} on the
sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
Each debugging scenario had a timeout limit of 4 minutes and a
memory limit of 6GB. The complete experiment required
over 30,000 hours of compute hours.

>>>>

\input{mutate-benchmarks-figure.tex}
\sub{benchmarks} {Which Software to Use As Benchmarks}

\subsection{XXX}
We run the experimental process from section~\ref{sec:rational} on the
sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
Each debugging scenario had a timeout limit of 4 minutes and a
memory limit of 6GB. The complete experiment required
over 30,000 hours of compute hours.
